### 2021-06-03 papers/thesis/full/ch3-disk-slices/disk-slices.tex
Further edits. New spy plots to make uniform across thesis. New references and fixing of capitals etc.

In recent years, there has been some interesting work in the area of multivariate orthogonal polynomials in two-dimensional domains (see \cite{dunkl2014orthogonal}), such as for a Chebychev-based spectral method on the rectangle \cite{julien2009efficient} \bstodo{list some domains and cite references for them}. In particular, spectral methods have been established on the disk \cite{boyd2011comparing, vasil2016tensor} and the triangle \cite{olver2019triangle}. Jacobi polynomials are extremely useful for solving a number of problems in not just 1D but 2D also, since one can utilise the sparsity of the relations that they hold to yield operator matrices that are efficient to solve \cite{doha2006efficient}. Applications for the disk include \bstodo{give applications and references} and for the triangle include solving Volterra integral equations \cite{gutleb2020sparse}, \bstodo{give more}.

In a similar vein to the ultraspherical method for the unit interval \cite{olver2013fast, doha2009efficient}, for the triangle a choice of OPs are the Jacobi polynomials on the triangle, which are a three-parameter family of polynomials in $(x,y)$ with differential operators acting in such a way so as to increment or decrement the parameter values \cite{dunkl2014orthogonal, koornwinder1975two, olver2019triangle}:
\bseqnnumber{
	P^{(a,b,c)}_{n,k}(x,y) := \tilde P^{(2k+b+c+1,a)}_{n-k}(x) \; (1-x)^k \; \tilde P^{(c,b)}_k\fpr({y \over 1-x}) \label{eqn:triangleops}
}
for $(x,y) \in \bsset{(x,y) \in \R^2}{0 < x < 1, \: 0 < y < 1-x}$, where $\tilde P_k^{(a,b)}$ is the degree $k$ \textit{shifted Jacobi polynomial}\footnote{The shifted Jacobi polynomials are orthogonal on the interval $[0,1]$ and given by $\tilde P_k^{(a,b)}(x) = P_k^{(a,b)}(2x - 1)$ where $P_k^{(a,b)}$ is the standard degree $k$ Jacobi polynomial.} with $a,b,c$ being parameter values.

For the disk, a review of spectral methods on the disk for solving the Poisson equation has been completed \cite{boyd2011comparing}, that looks at methods involving various bases for expanding functions in the disk including Zernike polynomials and Chebyshev polynomials. It is concluded that using Chebyshev polynomials yields mixed results, but that there is also no single advised basis to use. While Zernike polynomials yield accurate results, there is a lack of radial fast transform, which is a disadvantage. However, ideas for how to overcome this have progressed \cite{slevinsky2018use, sakai2009application, wilber2017computing}. One solution proposed is a radial basis defined in terms of Jacobi polynomials \cite{vasil2016tensor}:
\bseqnnumber{
	Q_n^{k,m}(r) := C_n^{k,m} r^m P^{(k,m)}_n(2r^2 - 1), \label{eqn:diskops}.
}
where $C_n^{k,m}$ is some normalising constant so that they are orthonormal with respect to $\int_0^1 Q_n^{k,m}(r) \; Q_{n'}^{k,m}(r) \; (1-r^2)^k \; r \; \D r = \delta_{n,n'}$. $\{Q_n^{k,m}\}$ are therefore a two-parameter family of orthogonal polynomials. Again, the idea here is that differential and other operators, as well as basis (parameter) conversion operators, can be represented as banded matrices.

In this chapter, we aim to present a basis and develop a sparse spectral method on generalisations of these domains, namely the disk-slice and trapezium. The full unit disk and triangle cases are special cases of the spectral methods we investigate here too. We will show that the operators for multiplication by $x$ and $y$, as well as differential and parameter conversion, will be banded-block-banded.

@@

We will show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations on the interval \cite{olver2013fast}, and its analogues on the disk \cite{vasil2016tensor} and triangle \cite{olver2018recurrence,olver2019triangle}. On the disk-slice the family of weights we consider are of the form
\bseqn{
	\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c
}
for $\alpha \leq x \leq \beta,$ $-\rho(x) \leq y \leq \rho(x)$. The corresponding OPs denoted $\hdopnkabc(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
\bseqn{
	\hdopnkabc(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
}
where $C_{n,k} \neq 0$ and \enquote{lower order terms'} includes degree $n$ polynomials of the form $x^{n-j} y^j$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs we will see in \bsrefdef{def:OPconstruction}.

@@

\begin{proposition}[{\cite[p55--56]{dunkl2014orthogonal}}]\label{prop:construction}
	Let $w_1 : (\alpha,\beta) \: \to \R$, $w_2 : (\gamma,\delta) \: \to \R$ be weight functions with $\alpha,\beta,\gamma,\delta \in \R$, and let $\rho \: : \: (\alpha,\beta) \: \to (0,\infty)$ be such that either \bsrefcondition{cond:trap} or \bsrefcondition{cond:disk} with $w_2$ being an even function hold.
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then, for $n = 0,1,2,\dots$, $k = 0,\dots,n$, the 2D polynomials defined on $\Omega$ given by
\bseqn{
	\hdopnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k\fpr(\frac{y}{\rho(x)})
}
are orthogonal polynomials with respect to the weight $W(x,y) := w_1(x) \: w_2\fpr(\frac{y}{\rho(x)})$ on $\Omega$. 
\end{proposition}

For disk slices and trapeziums, we specialise Proposition \ref{prop:construction} in the following definitions. First we introduce notation for two families of univariate OPs.
\begin{definition}\label{def:OPconstruction}
	Let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\bseqn{
	\begin{cases}
		\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
		\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
	\end{cases}
}
and define the associated inner products by:
\bseqnnumber{
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ipjac}
}
where
\bseqnnumber{
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
}
Denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).
\end{definition}

We can now write down our 2D OP family.
\begin{definition}\label{def:constuction}
	For parameters $a,b,c,d \in \R$, define the four-parameter 2D orthogonal polynomials as the set $\{\hdopnk^{(a,b,c,d)}\}$ where:
\bseqn{
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
}
\end{definition}

@@

Further, note that $J_x^{(a,b,c,d)}, J_y^{(a,b,c,d)}$ are banded-block-banded matrices (see \bsrefdef{def:bandedblockbanded}). They are block-tridiagonal (block-bandwidths $(1,1)$):
\bseqn{
J_{x/y}^{(a,b,c,d)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
}
where the blocks themselves are diagonal for $J_x^{(a,b,c,d)}$ (sub-block-bandwidths $(0,0)$),
\bseqn{
A^x_n &:= \begin{pmatrix}
		\alphaabcd_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaabcd_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaabcd_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaabcd_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^\top \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber
}
and tridiagonal for $J_y^{(a,b,c,d)}$ (sub-block-bandwidths $(1,1)$),
\bseqn{
A^y_n &:= \begin{pmatrix}
		\betaabcd_{n,0,8} & \betaabcd_{n,0,9} & & & \\
		\betaabcd_{n,1,7}& \ddots & \ddots & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaabcd_{n,n,7} & \betaabcd_{n,n,8} & \betaabcd_{n,n,9} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		\betaabcd_{n,0,5} & \betaabcd_{n,0,6} & & \\
		\betaabcd_{n,1,4} & \ddots & \ddots & \\
		& \ddots & \ddots & \betaabcd_{n,n-1,6} \\
		& & \betaabcd_{n,n,4} & \betaabcd_{n,n,5}
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		\betaabcd_{n,0,2} & \betaabcd_{n,0,3} & & \\
		\betaabcd_{n,1,1} & \ddots & \ddots & \\
		& \ddots & \ddots & \betaabcd_{n,n-2,3} \\
		& & \ddots & \betaabcd_{n,n-1,2} \\
		& & & \betaabcd_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
}

@@

\begin{figure} 
\center
\begin{tikzpicture} 
% Box/square (4x4)
\draw[black,solid,ultra thin] (0,0)--(0,4);
\draw[black,solid,ultra thin] (0,0)--(4,0);
\draw[black,solid,ultra thin] (4,0)--(4,4);
\draw[black,solid,ultra thin] (0,4)--(4,4);
% Dots
\draw[black,fill=black] (0,0) circle (.5ex);
\draw[black,fill=black] (2,2) circle (.5ex);
\draw[black,fill=black] (4,4) circle (.5ex);
\draw[black,fill=black] (0,2) circle (.5ex);
\draw[black,fill=black] (0,4) circle (.5ex);
% Arrows
\draw[black,thick,->] (0.2,0.2)--(1.8,1.8);
\draw[black,thick,->] (2.2,2.2)--(3.8,3.8);
\draw[black,thick,->] (0,0.2)--(0,1.8);
\draw[black,thick,->] (0,2.2)--(0,3.8);
\draw [black,thick,->] (0.2,4.2) to [out=20,in=160] (3.8,4.2);
% Node (parameter) labels
\draw[] (0,0) node[anchor=east] {$(0,0,0)$};
\draw[] (0,4) node[anchor=east] {$(0,0,2)$};
\draw[] (4,0) node[anchor=west] {$(2,2,0)$};
\draw[] (4,4) node[anchor=west] {$(2,2,2)$};
% Arrow labels
\draw[] (1,1) node[anchor=north] {$\tfrac{\partial}{\partial x}$};
\draw[] (3,3) node[anchor=north] {$\tfrac{\partial}{\partial x}$};
\draw[] (0,1) node[anchor=west] {$\tfrac{\partial}{\partial y}$};
\draw[] (0,3) node[anchor=west] {$\tfrac{\partial}{\partial y}$};
% The key and headings
\draw[black,thick,->] (-2.5,3.4)--(-2.5,4);
\draw[black,thick,->] (-2.5,3.4)--(-1.9,3.4);
\draw[] (-2.3,3.2) node[anchor=west] {$(a,b)$};
\draw[] (-2.7,3.6) node[anchor=south] {$c$};
\draw[] (2,4.6) node[anchor=south] {basis conversion};
\end{tikzpicture} 
\caption{The Laplace operator acting on vectors of $\smash{\hdopnk^{(0,0,0)}}$ coefficients has a sparse matrix representation if the range is represented as vectors of $\smash{\hdopnk^{(2,2,2)}}$ coefficients. Here, the arrows indicate that the corresponding operation has a sparse matrix representation when the domain is $\smash{\hdopnkabc}$ coefficients, where $(a,b,c)$ is at the tail of the arrow, and the range is $\smash{\hdopnk^{(\tilde{a},\tilde{b},\tilde{c})}}$ coefficients, where $(\tilde{a},\tilde{b},\tilde{c})$ is at the head of the arrow. %For example, $\smash{\frac{\partial}{\partial x}}$ has a sparse matrix representation when the domain is represented in $\smash{\hdop_{n,k}^{(0,0,0)}}$ (resp.~$\smash{\hdop_{n,k}^{(1,0,1)}}$) and the range in $\smash{\hdop_{n,k}^{(1,0,1)}}$ (resp.~$\smash{\hdop_{n,k}^{(2,0,2)}}$).
}
\label{fig:laplaceparameterconversion} 
\end{figure}

@@

\begin{definition}\label{def:differentialoperators}
Define the operator matrices $D_x^{(a,b,c)}, \: D_y^{(a,b,c)}, \: W_x^{(a,b,c)}, \: W_y^{(a,b,c)}$ according to:
\bseqn{
{\partial f \over \partial x}&= \bighdop^{(a+1,b+1,c+1)}(x,y)^\top \: D_x^{(a,b,c)} \: \mathbf{f}, \\
{\partial f \over \partial y} &= \bighdop^{(a,b,c+1)}(x,y)^\top \: D_y^{(a,b,c)} \: \mathbf{f}, \\
{\partial \over \partial x}[\Wabc(x,y) \: f(x,y)] &= \bigW^{(a-1,b-1,c-1)}(x,y)^\top \: W_x^{(a,b,c)} \: \mathbf{f}, \\
{\partial \over \partial y}[\Wabc(x,y) \: f(x,y)] &= \bigW^{(a,b,c-1)}(x,y)^\top \: W_y^{(a,b,c)} \: \mathbf{f}.
}
\end{definition}

The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomial families' derivatives, for example the Jacobi polynomials on the interval, as seen in the DLMF \cite[(18.9.3)]{DLMF}, and on the triangle \cite{olver2018recurrence}. An illustration of how the non-weighted differential operators increment the parameters (a,b,c) is seen in Figure \ref{fig:laplaceparameterconversion}.

\begin{theorem}\label{theorem:sparsityofdifferentialoperators}
The operator matrices $D_x^{(a,b,c)}, \: D_y^{(a,b,c)}, \: W_x^{(a,b,c)}, \: W_y^{(a,b,c)}$ from Definition \ref{def:differentialoperators} are sparse, with banded-block-banded structure. More specifically:
\begin{itemize}
	\item $D_x^{(a,b,c)}$ has  block-bandwidths $(-1,3)$, and sub-block-bandwidths $(0, 2)$.
  	\item $D_y^{(a,b,c)}$ has  block-bandwidths $(-1,1)$, and sub-block-bandwidths $(-1,1)$.
	\item $W_x^{(a,b,c)}$ has  block-bandwidths $(3,-1)$, and sub-block-bandwidths $(2, 0)$.
  	\item $W_y^{(a,b,c)}$ has  block-bandwidths $(1,-1)$, and sub-block-bandwidths $(1,-1)$.
\end{itemize}
\end{theorem}

@@

The sparsity of operator matrices for partial differentiation by $x, y$ as well as for parameter transformations generalise to such end-disk-slice domains. For instance, if we inspect the proof of Lemma \ref{theorem:sparsityofdifferentialoperators}, we see that it can easily generalise to the weights and domain $\Omega$ for an end-disk-slice.

>>>>

In recent years, there has been some interesting work in the area of multivariate orthogonal polynomials in two-dimensional domains (see \cite{dunkl2014orthogonal}), such as for a Chebychev-based spectral method on the rectangle \cite{julien2009efficient}. In particular, spectral methods have been established on the disk \cite{boyd2011comparing, vasil2016tensor} and the triangle \cite{olver2019triangle}. Examples of applications for the disk include use in optics \cite{mahajan2007orthonormal}, astrophysics \cite{pringle1981accretion}, and fluid dynamics \cite{eggels1994fully, noll1976zernike, kerswell2005recent}, and for the triangle include solving Volterra integral equations \cite{gutleb2020sparse}.

For the disk, a review of spectral methods on the disk for solving the Poisson equation has been completed \cite{boyd2011comparing}, that looks at methods involving various bases for expanding functions in the disk including Zernike polynomials and Chebyshev polynomials. It is concluded that using Chebyshev polynomials yields mixed results, but that there is also no single advised basis to use. While Zernike polynomials yield accurate results, there is a lack of radial fast transform, which is a disadvantage. However, ideas for how to overcome this have progressed \cite{slevinsky2018use, sakai2009application, wilber2017computing}. One solution proposed is pairing the Zernike disk OPs with a radial basis defined in terms of Jacobi polynomials \cite{vasil2016tensor}:
\bseqnnumber{
	Q_n^{k,m}(r) := C_n^{k,m} r^m P^{(k,m)}_n(2r^2 - 1), \label{eqn:diskops}.
}
where $C_n^{k,m}$ is some normalising constant so that they are orthonormal with respect to $\int_0^1 Q_n^{k,m}(r) \; Q_{n'}^{k,m}(r) \; (1-r^2)^k \; r \; \D r = \delta_{n,n'}$. $\{Q_n^{k,m}\}$ are therefore a two-parameter family of orthogonal polynomials. The use of polar coordinates here then allows for radially symmetric partial differential (and other) operators, as well as basis (parameter) conversion operators, to be represented as banded matrices by exploiting the relations of the Jacobi polynomials -- they are extremely useful for solving a number of problems in not just 1D but 2D also, since one can utilise the sparsity of the relations that they hold to yield operator matrices that are efficient to solve \cite{doha2006efficient}.

Conversely, in a similar vein to the ultraspherical method for the unit interval \cite{olver2013fast, doha2009efficient}, a choice of OPs for the triangle are the \textit{Jacobi polynomials on the triangle}, which are a three-parameter family of polynomials in $(x,y)$ with differential operators acting in such a way so as to increment or decrement the parameter values \cite{dunkl2014orthogonal, koornwinder1975two, olver2019triangle}:
\bseqnnumber{
	P^{(a,b,c)}_{n,k}(x,y) := \tilde P^{(2k+b+c+1,a)}_{n-k}(x) \; (1-x)^k \; \tilde P^{(c,b)}_k\fpr({y \over 1-x}) \label{eqn:triangleops}
}
for $(x,y) \in \bsset{(x,y) \in \R^2}{0 < x < 1, \: 0 < y < 1-x}$, where $\tilde P_k^{(a,b)}$ is the degree $k$ \textit{shifted Jacobi polynomial}\footnote{The shifted Jacobi polynomials are orthogonal on the interval $[0,1]$ and given by $\tilde P_k^{(a,b)}(x) = P_k^{(a,b)}(2x - 1)$ where $P_k^{(a,b)}$ is the standard degree $k$ Jacobi polynomial.} with $a,b,c$ being parameter values. Since there is no radial symmetry, differential operators for the triangle (as will be the case for the domains we investigate in this chapter) will be more complex than those for the disk.

In this chapter, we aim to present a basis and develop a sparse spectral method on generalisations of these domains, namely the disk-slice and trapezium. The full unit disk and triangle cases can be viewed as special cases of the spectral methods we investigate here too. We will show that the operators for multiplication by $x$ and $y$, as well as differential and parameter conversion, will be banded-block-banded.

@@

We will show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations on the interval \cite{olver2013fast}, and its analogues on the disk \cite{vasil2016tensor} and triangle \cite{olver2018recurrence,olver2019triangle}. On the disk-slice the family of weights we consider are of the form
\bseqn{
	\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c
}
for $\alpha \leq x \leq \beta,$ $-\rho(x) \leq y \leq \rho(x)$. The corresponding OPs denoted $\hdopnkabc(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
\bseqn{
	\hdopnkabc(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
}
where $C_{n,k} \neq 0$ and \enquote{lower order terms'} includes degree $n$ polynomials of the form $x^{n-j} y^j$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs we will see in \bsrefdef{def:ds:OPconstruction}.

@@

\begin{proposition}[{\cite[p55--56]{dunkl2014orthogonal}}]\label{prop:ds:construction}
	Let $w_1 : (\alpha,\beta) \: \to \R$, $w_2 : (\gamma,\delta) \: \to \R$ be weight functions with $\alpha,\beta,\gamma,\delta \in \R$, and let $\rho \: : \: (\alpha,\beta) \: \to (0,\infty)$ be such that either \bsrefcondition{cond:trap} or \bsrefcondition{cond:disk} with $w_2$ being an even function hold.
$\forall$, $n = 0,1,2,\dots, $ let $\{p_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k+1} w_1(x)$ where $0 \le k \le n$, and $\{q_{n}\}$ be polynomials orthogonal with respect to the weight $w_2(x)$. Then, for $n = 0,1,2,\dots$, $k = 0,\dots,n$, the 2D polynomials defined on $\Omega$ given by
\bseqn{
	\hdopnk(x,y) := p_{n-k,k}(x) \: \rho(x)^k \: q_k\fpr(\frac{y}{\rho(x)})
}
are orthogonal polynomials with respect to the weight $W(x,y) := w_1(x) \: w_2\fpr(\frac{y}{\rho(x)})$ on $\Omega$. 
\end{proposition}

For disk slices and trapeziums, we specialise \bsrefprop{prop:ds:construction} in the following definitions. First we introduce notation for two families of univariate OPs.
\begin{definition}\label{def:ds:OPconstruction}
	Let $\genjacw^{(a,b,c)}(x)$ and $\jacw^{(a,b)}(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\bseqn{
	\begin{cases}
		\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
		\jacw^{(a,b)}(x) &:= (\delta-x)^a \: (x - \gamma)^b
	\end{cases}
}
and define the associated inner products by:
\bseqnnumber{
	\ip< p, \: q >_{\genjacw^{(a,b,c)}} &:= \frac{1}{\normgenjac^{(a,b,c)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b,c)}(x) \: \D x \label{eqn:ipgenjac} \\
	\ip<p, q>_{\jacw^{(a,b)}} &:= \frac{1}{\normjac^{(a,b)}} \: \int_\gamma^\delta p(y) \: q(y) \: \jacw^{(a,b)}(y)\: \D y \label{eqn:ipjac}
}
where
\bseqnnumber{
	\normgenjac^{(a,b,c)} := \int_\alpha^\beta \: \genjacw^{(a,b,c)}(x) \: \D x, \quad \normjac^{(a,b)} := \int_\gamma^\delta \: \jacw^{(a,b)}(y) \: \D y. \label{eqn:ipnormalisations}
}
Denote the three-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}), and the two-parameter family of orthonormal polynomials on $[\gamma,\delta]$ by $\{\jac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipjac}).
\end{definition}

We can now write down our 2D OP family.
\begin{definition}\label{def:ds:constuction}
	For parameters $a,b,c,d \in \R$, define the four-parameter 2D orthogonal polynomials as the set $\{\hdopnk^{(a,b,c,d)}\}$ where:
\bseqn{
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \jac_k^{(d,c)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
}
\end{definition}

@@

Further, note that $J_x^{(a,b,c,d)}, J_y^{(a,b,c,d)}$ are banded-block-banded matrices (see \bsrefdef{def:sh:bandedblockbanded}). They are block-tridiagonal (block-bandwidths $(1,1)$):
\bseqn{
J_{x/y}^{(a,b,c,d)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
}
where the blocks themselves are diagonal for $J_x^{(a,b,c,d)}$ (sub-block-bandwidths $(0,0)$),
\bseqn{
A^x_n &:= \begin{pmatrix}
		\alphaabcd_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaabcd_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaabcd_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaabcd_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^\top \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber
}
and tridiagonal for $J_y^{(a,b,c,d)}$ (sub-block-bandwidths $(1,1)$),
\bseqn{
A^y_n &:= \begin{pmatrix}
		\betaabcd_{n,0,8} & \betaabcd_{n,0,9} & & & \\
		\betaabcd_{n,1,7}& \ddots & \ddots & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaabcd_{n,n,7} & \betaabcd_{n,n,8} & \betaabcd_{n,n,9} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		\betaabcd_{n,0,5} & \betaabcd_{n,0,6} & & \\
		\betaabcd_{n,1,4} & \ddots & \ddots & \\
		& \ddots & \ddots & \betaabcd_{n,n-1,6} \\
		& & \betaabcd_{n,n,4} & \betaabcd_{n,n,5}
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		\betaabcd_{n,0,2} & \betaabcd_{n,0,3} & & \\
		\betaabcd_{n,1,1} & \ddots & \ddots & \\
		& \ddots & \ddots & \betaabcd_{n,n-2,3} \\
		& & \ddots & \betaabcd_{n,n-1,2} \\
		& & & \betaabcd_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
}

@@

\begin{figure} 
\center
\begin{tikzpicture} 
% Box/square (4x4)
\draw[black,solid,ultra thin] (0,0)--(0,4);
\draw[black,solid,ultra thin] (0,0)--(4,0);
\draw[black,solid,ultra thin] (4,0)--(4,4);
\draw[black,solid,ultra thin] (0,4)--(4,4);
% Dots
\draw[black,fill=black] (0,0) circle (.5ex);
\draw[black,fill=black] (2,2) circle (.5ex);
\draw[black,fill=black] (4,4) circle (.5ex);
\draw[black,fill=black] (0,2) circle (.5ex);
\draw[black,fill=black] (0,4) circle (.5ex);
% Arrows
\draw[black,thick,->] (0.2,0.2)--(1.8,1.8);
\draw[black,thick,->] (2.2,2.2)--(3.8,3.8);
\draw[black,thick,->] (0,0.2)--(0,1.8);
\draw[black,thick,->] (0,2.2)--(0,3.8);
\draw [black,thick,->] (0.2,4.2) to [out=20,in=160] (3.8,4.2);
% Node (parameter) labels
\draw[] (0,0) node[anchor=east] {$(0,0,0)$};
\draw[] (0,4) node[anchor=east] {$(0,0,2)$};
\draw[] (4,0) node[anchor=west] {$(2,2,0)$};
\draw[] (4,4) node[anchor=west] {$(2,2,2)$};
% Arrow labels
\draw[] (1,1) node[anchor=north] {$\tfrac{\partial}{\partial x}$};
\draw[] (3,3) node[anchor=north] {$\tfrac{\partial}{\partial x}$};
\draw[] (0,1) node[anchor=west] {$\tfrac{\partial}{\partial y}$};
\draw[] (0,3) node[anchor=west] {$\tfrac{\partial}{\partial y}$};
% The key and headings
\draw[black,thick,->] (-2.5,3.4)--(-2.5,4);
\draw[black,thick,->] (-2.5,3.4)--(-1.9,3.4);
\draw[] (-2.3,3.2) node[anchor=west] {$(a,b)$};
\draw[] (-2.7,3.6) node[anchor=south] {$c$};
\draw[] (2,4.6) node[anchor=south] {basis conversion};
\end{tikzpicture} 
\caption{The Laplace operator acting on vectors of $\smash{\hdopnk^{(0,0,0)}}$ coefficients has a sparse matrix representation if the range is represented as vectors of $\smash{\hdopnk^{(2,2,2)}}$ coefficients. Here, the arrows indicate that the corresponding operation has a sparse matrix representation when the domain is $\smash{\hdopnkabc}$ coefficients, where $(a,b,c)$ is at the tail of the arrow, and the range is $\smash{\hdopnk^{(\tilde{a},\tilde{b},\tilde{c})}}$ coefficients, where $(\tilde{a},\tilde{b},\tilde{c})$ is at the head of the arrow. %For example, $\smash{\frac{\partial}{\partial x}}$ has a sparse matrix representation when the domain is represented in $\smash{\hdop_{n,k}^{(0,0,0)}}$ (resp.~$\smash{\hdop_{n,k}^{(1,0,1)}}$) and the range in $\smash{\hdop_{n,k}^{(1,0,1)}}$ (resp.~$\smash{\hdop_{n,k}^{(2,0,2)}}$).
}
\label{fig:ds:laplaceparameterconversion} 
\end{figure}

@@

\begin{definition}\label{def:ds:differentialoperators}
Define the operator matrices $D_x^{(a,b,c)}, \: D_y^{(a,b,c)}, \: W_x^{(a,b,c)}, \: W_y^{(a,b,c)}$ according to:
\bseqn{
{\partial f \over \partial x}&= \bighdop^{(a+1,b+1,c+1)}(x,y)^\top \: D_x^{(a,b,c)} \: \mathbf{f}, \\
{\partial f \over \partial y} &= \bighdop^{(a,b,c+1)}(x,y)^\top \: D_y^{(a,b,c)} \: \mathbf{f}, \\
{\partial \over \partial x}[\Wabc(x,y) \: f(x,y)] &= \bigW^{(a-1,b-1,c-1)}(x,y)^\top \: W_x^{(a,b,c)} \: \mathbf{f}, \\
{\partial \over \partial y}[\Wabc(x,y) \: f(x,y)] &= \bigW^{(a,b,c-1)}(x,y)^\top \: W_y^{(a,b,c)} \: \mathbf{f}.
}
\end{definition}

The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomial families' derivatives, for example the Jacobi polynomials on the interval, as seen in the DLMF \cite[(18.9.3)]{DLMF}, and on the triangle \cite{olver2018recurrence}. An illustration of how the non-weighted differential operators increment the parameters (a,b,c) is seen in \bsreffig{fig:ds:laplaceparameterconversion}.

\begin{theorem}\label{theorem:ds:sparsityofdifferentialoperators}
The operator matrices $D_x^{(a,b,c)}, \: D_y^{(a,b,c)}, \: W_x^{(a,b,c)}, \: W_y^{(a,b,c)}$ from \bsrefdef{def:ds:differentialoperators} are sparse, with banded-block-banded structure. More specifically:
\begin{itemize}
	\item $D_x^{(a,b,c)}$ has  block-bandwidths $(-1,3)$, and sub-block-bandwidths $(0, 2)$.
  	\item $D_y^{(a,b,c)}$ has  block-bandwidths $(-1,1)$, and sub-block-bandwidths $(-1,1)$.
	\item $W_x^{(a,b,c)}$ has  block-bandwidths $(3,-1)$, and sub-block-bandwidths $(2, 0)$.
  	\item $W_y^{(a,b,c)}$ has  block-bandwidths $(1,-1)$, and sub-block-bandwidths $(1,-1)$.
\end{itemize}
\end{theorem}

@@

The sparsity of operator matrices for partial differentiation by $x, y$ as well as for parameter transformations generalise to such end-disk-slice domains. For instance, if we inspect the proof of \bsreftheorem{theorem:ds:sparsityofdifferentialoperators}, we see that it can easily generalise to the weights and domain $\Omega$ for an end-disk-slice.
